True versus apparent shapes of bow shocks and bow waves

### 2017-11-15 papers/Paper1/app-dust-wave.tex
Add figure with on-axis grain trajectories

\newcommand\Qp{\ensuremath{Q_{\text{p}}}}
\newcommand{\grain}{\ensuremath{_{\text{g}}}}
\newcommand{\xsec}{\ensuremath{\sigma\grain}}
\newcommand\frad{\ensuremath{f_{\text{rad}}}}
\newcommand\thm{\ensuremath{\theta_{\text{m}}}}
A dust grain of geometrical cross-section \(\xsec\) situated a
distance \(R\) from a point source of radiation with luminosity \(L\)
will experience a repulsive, radially directed radiative force
\citep[e.g.,][]{Spitzer:1978a}
\begin{equation}
  \label{eq:dust-rad-force}
  \frad = \frac{\xsec \Qp L} {4 \pi R^2 c} e^{-\tau}
\end{equation}
where \Qp{} is the frequency-averaged\footnote{%
  Frequency averages of any quantity \(x\) should be understood as
  weighted by the attenuated source spectrum:
  \(\langle x \rangle_\nu = (L \, e^{-\tau})^{-1} \int_0^\infty x(\nu)\, L_\nu \, e^{-\tau_\nu} \, d\nu
  \).  } %
radiation pressure efficiency\footnote{%
  For absorption efficiency \(Q_{\text{abs}}\), scattering efficiency
  \(Q_{\text{scat}}\), and asymmetry parameter (mean scattering
  cosine) \(g\), we have
  \(\Qp = Q_{\text{abs}} + (1 - g) Q_{\text{scat}}\)
  \citep[e.g., \S~4.5 of][]{Bohren:1983a}.} %
of the grain, \(c\) is the speed of light, and \(\tau\) is the
frequency-averaged optical depth between the source and the grain.
For simplicity, we will consider only the optically thin case,
\(\tau \to 0\).

@@

The length scale, \(R_0\), of the bow wave depends on the grain
properties via the combination \(\xsec \Qp / m\grain\).  For grains of
size \(a\grain\) and internal density \(\rho\grain\), we have
\(\xsec / m\grain \approx (a\grain \rho\grain)^{-1}\).  For radiation with
wavelength smaller than the grain size, \(\lambda < a\grain\), the
efficiency is \(\Qp \sim 1\), whereas for \(\lambda > a\grain\) it is
\(\Qp \sim a\grain/\lambda\).  Therefore, we would expect \(R_0\) to be almost
independent of grain size for small grains, but to vary as
\(R_0 \propto a\grain^{-1}\) for large grains, where small/large is relative
to the peak wavelength of the radiation source.

@@

>>>>

\newcommand\Qp{\ensuremath{Q_{\text{p}}}}
\newcommand{\grain}{\ensuremath{_{\text{d}}}}
\newcommand{\xsec}{\ensuremath{\sigma\grain}}
\newcommand\frad{\ensuremath{f_{\text{rad}}}}
\newcommand\thm{\ensuremath{\theta_{\text{m}}}}
A dust grain of geometrical cross-section \(\xsec\) situated a
distance \(R\) from a point source of radiation with luminosity \(L\)
will experience a repulsive, radially directed radiative force
\citep[e.g.,][]{Spitzer:1978a}
\begin{equation}
  \label{eq:dust-rad-force}
  \frad = \frac{\xsec \Qp L} {4 \pi R^2 c} e^{-\tau}
\end{equation}
where \Qp{} is the frequency-averaged\footnote{%
  Frequency averages of any quantity \(x\) should be understood as
  weighted by the attenuated source spectrum:
  \(\langle x \rangle_\nu = (L \, e^{-\tau})^{-1} \int_0^\infty x(\nu)\, L_\nu \, e^{-\tau_\nu} \, d\nu
  \).  } %
radiation pressure efficiency\footnote{%
  For absorption efficiency \(Q_{\text{abs}}\), scattering efficiency
  \(Q_{\text{scat}}\), and asymmetry parameter (mean scattering
  cosine) \(g\), we have
  \(\Qp = Q_{\text{abs}} + (1 - g) Q_{\text{scat}}\)
  \citep[e.g., \S~4.5 of][]{Bohren:1983a}.} %
of the grain, \(c\) is the speed of light, and \(\tau\) is the
frequency-averaged optical depth between the source and the grain.
For simplicity, we will consider only the optically thin case,
\(\tau \to 0\).

\subsection{Gas-free bow wave}
\label{sec:gas-free-bow}

@@

The apex stagnation radius, \(R_0\), of the bow wave depends on the
grain properties via the combination \(\xsec \Qp / m\grain\).  For
grains of size \(a\grain\) and internal density \(\rho\grain\), we have
\(\xsec / m\grain \approx (a\grain \rho\grain)^{-1}\).  For radiation with
wavelength smaller than the grain size, \(\lambda < a\grain\), the
efficiency is \(\Qp \sim 1\), whereas for \(\lambda > a\grain\) it is
\(\Qp \sim a\grain/\lambda\).  Therefore, we would expect \(R_0\) to be almost
independent of grain size for small grains, but to vary as
\(R_0 \propto a\grain^{-1}\) for large grains, where small/large is relative
to the peak wavelength of the radiation source.  In principle, a
polydisperse population of grains could produce a blurring of the
observed bow wave, but only if large grains contribute significantly
to the dust emission.

\newcommand\drag{\ensuremath{_{\text{drag}}}}
\newcommand{\gas}{\ensuremath{_{\text{gas}}}}
\newcommand{\drift}{\ensuremath{_{\text{drift}}}}
\newcommand\soundspeed{\ensuremath{c_{\text{s,gas}}}}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/dust-coupling-1d}
  \caption{Dust-gas coupling for an on-axis (purely radial) trajectory
    with \(\alpha\drag = 0.5\). (a)~Grain velocity (in units of
    \(v_\infty\)) and radius (in units of \(R_0\)) versus time. (b)~Phase
    space trajectory. Grain spirals in on point
    \((x, u) = (\alpha\drag^{-1}, 0)\).}
  \label{fig:dust-coupling-1d}
\end{figure}

@@

\subsection{Bow wave with gas drag}
\label{sec:bow-wave-drag}

More realistically, a grain will also be subject to a drag force,
\(f\drag\), due to its relative motion with respect to gas or plasma
particles. If the gas density, velocity, and sound speed are
\(\rho\gas\), \(v\gas\), and \(\soundspeed\), then a grain with velocity
\(v\grain\) will experience a drag force that is directed opposite to
the relative velocity, \(w = v\grain - v\gas\).  In the supersonic
limit, \(\abs{w} \gg \soundspeed\), the magnitude of the force is
\begin{equation}
  \label{eq:dust-fdrag}
  \abs{f\drag} \approx Q\drag \xsec \rho\gas w^2 \ ,
\end{equation}
where \(Q\drag\) is a boost factor to account for the increase in
cross section due to the Coulomb force when a charged grain interacts
with an ionized plasma \citep{Draine:1979a}.  We neglect the back
reaction of the dust on the gas motion and assume a uniform background
gas flow with \(v\gas = -v_\infty\).

Considering the incoming flow on the symmetry axis, at each radius
there is an asymptotic gas--grain drift speed, \(w\drift\), for which
the radiative and drag forces exactly cancel, \(f\drag = -\frad\),
yielding
\begin{equation}
  \label{eq:dust-wdrift}
  w\drift = \left( \frac{\Qp L} {4\pi c Q\drag \rho\gas R^2} \right)^{1/2} \ .
\end{equation}
Any deviation of \(w\) from \(w\drift\) produces unbalanced forces
that tend to restore \(w \to w\drift\).  We define a dimensionless
coupling coefficient, \(\alpha\drag\), to be the speed of the incoming
stream in units of the drift velocity at the radiative stagnation
radius:
\begin{equation}
  \label{eq:dust-alpha}
  \alpha\drag \equiv \frac{v_\infty} {w\drift(R_0)} = \left(
    Q\drag \frac{R_0 / a\grain} {\rho\grain / \rho\gas}
  \right)^{1/2} \ ,
\end{equation}
where we have used equation~\eqref{eq:dust-r0} and suppressed a
grain-shape dependent geometric factor of order unity.  If
\(\alpha\drag \ll 1\), then \(w\drift \gg v_\infty\) out to several times the
stagnation radius, so the radiation field has no difficulty in
effectively decoupling the grain from the gas and producing the
velocity difference, \(w = 2 v_\infty\), that is required to turn the grain
around and expel it towards the direction whence it came.  However,
for non-zero \(\alpha\drag\) the \(R^{-1}\) dependence of \(w\drift\)
(eq.~\eqref{eq:dust-wdrift}) means that the grain will
\textit{re-couple} to the inflowing gas stream around a radius
\(\approx R_0 / \alpha\drag \) and be swept in again for another try.  This
behavior is illustrated in Figure~\ref{fig:dust-coupling-1d}, where it can be seen that
